---
title: "Multivariate Project 1"
author: "Io"
date: "2025-04-23"
output: html_document
---

## Introduction

This report explores the relationship between education and perceptions of climate change, using data from the European Social Survey (ESS Round 11). In particular, a comparative analysis is performed on two countries: Norway and Portugal.

#### Loading Packages

```{r}
library(ggplot2)
library(dplyr)
library(readr)
library(knitr)

```

#### Loading the dataset

```{r}
ESS11 <- read_csv("data/ESS11.csv")
```

## Exploratory data analysis

Some preliminary steps are taken in order to clean the data from NAs and to visualize variable distributions and some basic bivariate relationships.

#### Preparing the data

Beliefs about the causes of climate change - natural vs. human:

```{r}
ESS11$ccnthum[ESS11$ccnthum %in% c(55, 77, 88, 99)] <- NA

clim_change_cause <- factor(ESS11$ccnthum,
                        levels = c(1, 2, 3, 4, 5),
                        labels = c("Entirely natural", "Mainly natural", "Equally natural and human", "Mainly human", "Entirely human"))

```

Beliefs about individual responsibility in fighting climate change:

```{r}
ESS11$ccrdprs[ESS11$ccrdprs %in% c(66, 77, 88, 99)] <- NA

ESS11 <- ESS11 %>%
  mutate(
    clim_change_resp = case_when(
      ccrdprs >= 1 & ccrdprs <= 3 ~ "Low",
      ccrdprs >= 4 & ccrdprs <= 7 ~ "Medium",
      ccrdprs >= 8 & ccrdprs <= 10 ~ "High"
    ),
    clim_change_resp = factor(clim_change_resp, levels = c("Low", "Medium", "High"), ordered = TRUE)
  )
```

Education
```{r}

ESS11$edulvlb_clean <- NA

ESS11$edulvlb_clean[ESS11$edulvlb %in% c(113, 129)] <- 1  
ESS11$edulvlb_clean[ESS11$edulvlb %in% c(131:223)] <- 2  
ESS11$edulvlb_clean[ESS11$edulvlb %in% c(321:323, 412, 413)] <- 3 
ESS11$edulvlb_clean[ESS11$edulvlb %in% c(421:423)] <- 4  
ESS11$edulvlb_clean[ESS11$edulvlb %in% c(510:520)] <- 5 
ESS11$edulvlb_clean[ESS11$edulvlb %in% c(610, 620)] <- 6 
ESS11$edulvlb_clean[ESS11$edulvlb == 800] <- 7  

ESS11_clean <- ESS11[!is.na(ESS11$edulvlb_clean), ]

```



#### Descriptive statistics

Frequency tables are employed in order to better understand the distribution of the variables.

###### Climate change and perceived causes

```{r}
summary(ESS11$clim_change_cause)

prop_table <- prop.table(table(clim_change_cause))

table_out <- data.frame(
  Cause = names(prop_table),
  Percentage = round(as.numeric(prop_table) * 100, 1)
)
kable(table_out, caption = "Beliefs about causes of climate change (in %)")

```

###### Climate change and individual responsibility

```{r}
summary(ESS11$clim_change_resp)

prop_table <- prop.table(table(ESS11$clim_change_resp))

table_out <- data.frame(
  Responsibility = names(prop_table),
  Percentage = round(as.numeric(prop_table) * 100, 1)
)

kable(table_out, caption = "Perceived personal responsibility for reducing climate change (in %)")

```

###### Education
```{r}
education <- ESS11_clean %>%
  group_by(cntry) %>%
  summarise(media_edulvlb = mean(edulvlb_clean)) %>%
  arrange(desc(media_edulvlb))

kable(education, caption = "Average education level (by country)")

```



#### Plots

Bar graphs allow for the visualization of the univariate distribution of the variables associated with beliefs about climate change.

```{r}
ggplot(ESS11, aes(x = clim_change_cause)) +
  geom_bar(color = "purple", fill = "purple") +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
         axis.text.y = element_text(size = 12)) +
  labs(title = "Beliefs about causes of climate change",
       x = "Natural vs. human",
       y = "Count")


ESS11 |> 
  filter(!is.na(clim_change_resp)) |>  #I want to exclude NAs from the bar graph
  ggplot(aes(x = clim_change_resp)) +
  geom_bar(color = "purple", fill = "purple") +
  theme(axis.text.x = element_text(size = 12, hjust = 1),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold")) +
  labs(title = "Feelings of personal responsibilities to reduce climate change",
       x = "Perceived responsibility",
       y = "Count")


ggplot(education, aes(x = reorder(cntry, media_edulvlb), y = media_edulvlb)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Average of higher level of education (ESS11)",
       x = "Country", y = "Average level of education (1 = low, 7 = high)") +
  theme_minimal()

```

#### Comparative analysis

Filtering for countries of interest (Norway and Portugal).

```{r}
ESS11_NO_PT <- ESS11_clean %>%
  filter(cntry %in% c("NO", "PT"))
```

##### Beliefs about personal responsibility towards climate change, by education.

###### Differences in levels of perceived responsibility in Norway and Portugal
```{r}
ggplot(ESS11_NO_PT, aes(x = clim_change_resp, fill = cntry)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("NO" = "steelblue", "PT" = "tomato")) +
  labs(
    title = "Perceived responsibility for climate change: Norway vs Portugal",
    x = "Perceived responsibility",
    y = "Count",
    fill = "Country"
  ) +
  theme_minimal()
```

###### Regression model
```{r}
model_NO <- lm(as.numeric(clim_change_resp) ~ edulvlb_clean, data = ESS11_NO_PT %>% filter(cntry == "NO"))
model_PT <- lm(as.numeric(clim_change_resp) ~ edulvlb_clean, data = ESS11_NO_PT %>% filter(cntry == "PT"))

summary(model_NO)
summary(model_PT)
```


```{r}
ggplot(ESS11_NO_PT, aes(x = edulvlb_clean, y = as.numeric(clim_change_resp), color = cntry)) +
  geom_smooth(method = "lm", se = TRUE, size = 1.3) +
  scale_color_manual(values = c("NO" = "steelblue", "PT" = "tomato")) +
  labs(
    title = "Relationship between education and climate responsibility",
    subtitle = "Comparison between Norway and Portugal",
    x = "Education level (1 = low, 7 = high)",
    y = "Responsibility (1 = Low, 3 = High)",
    color = "Country"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

```



